HL-LHC Computing Review:\\Common Tools and Community Software

### 2020-04-30 LHCC/2020/2020-01/input_recoTrigger.tex
Add references and British English spelling changes

Experiments have also been working towards minimizing the differences
between trigger (online) and offline software for reconstruction and
calibration, so that more refined physics objects can be obtained
directly within the HLT farm for a more efficient event selection. This
is also in-line with enhancing the experiment's capabilities for
real-time data analysis of events accepted by the hardware trigger
system, driven by use cases where the physics potential improves when
analysing more events than what can be written out in full with
traditional data processing.

@@

The huge increase in the number of charged particles and hence the
combinatorial load for tracking algorithms at future colliders will put
great strain on compute resources. To minimize the memory footprint,
tracking software needs to be thread-safe to support multi-threaded
execution per core. At the same time, the software has to be efficient
and accurate to meet the physics requirements.

@@

The aim of the \textbf{mkFit} project is to speed up Kalman filter (KF)
tracking algorithms using highly parallel architectures and to deliver
track building (and possibly fitting) software for the HL-LHC. Recent
activities focused on delivering a production quality software setup to
perform track building in the context of the LHC Run-3 (or Run-2). The
implementation relies on single-precision floating point math and is
available for multicore CPUs and achieves significant gains in compute
performance from the use of vector instructions set extending to
AVX-512, based on Matriplex library (a part of the mkFIt project).

\textbf{Exa.TrkX}~\cite{Ju:2020xty} is a cross-experiment collaboration of data scientists
and computational physicists from ATLAS, CMS and DUNE. It develops
production-quality deep neural network models, in particular Graph
Neural Networks, for charged particle tracking on diverse detectors
employing next-generation computing architectures such as HPCs. It is
also exploring distributed training and optimization of Graph Neural
Networks (GNN) on HPCs, and the deployment of GNNs with microsecond
latencies on Level-1 trigger systems.

@@

\hypertarget{general-reconstruction-software-improvements-vectorization}{%
\subsubsection{General reconstruction software improvements:
vectorization}\label{general-reconstruction-software-improvements-vectorization}}

Improving the ability of HEP developed toolkits to use vector units on
commodity hardware will bring speedups to applications running on both
current and future hardware. The goal for work in this area is to evolve
current toolkit and algorithm implementations, and best programming
techniques to better use SIMD capabilities of current and future
computing architectures. Since the CWP, algorithm development projects
(e.g. \cite{LHCB-FIGURE-2019-002}
as well as mkFit) have demonstrated success in increasing the use of vector
units. In addition, HEP has increasingly benefited from industry
developed software, including machine learning toolkits (e.g.,
tensorflow), that typically make excellent use of vector units.
Challenges remain in this area. For example, full exploitation of this
must also account for many generations of hardware that experiments must
exploit on the grid. In addition, to realize full performance gains, a
large fraction of the application must be improved to use SIMD processor
capabilities (partly due to turbo-boost capabilities of processors)

@@

Work is already undergoing in the collaborations to evolve or rewrite
existing toolkits and algorithms focused on their physics and technical
performance at high event complexity (e.g. high pileup at HL-LHC), and
efforts in this area have expanded. Cross-collaboration developments are
focusing on the availability of ML algorithms for experimental software,
especially in persistifying models and running inferences in production
(e.g.
\emph{\href{https://github.com/lwtnn/lwtnn}{{https://github.com/lwtnn/lwtnn}}
and \href{https://onnx.ai}{{https://onnx.ai}}}) and in enhancing the
training capacities of collaborations by submitting jobs to the grid and
to facilities offering large CPU/GPU resources. One can also foresee
that HEP will increase the benefit from techniques and cross-talk from
industry.

@@

\textbf{Allen} is a data processing framework for GPUs, as well as a
specific implementation of a first-level trigger for LHCb Run 3
data-taking. Allen is optimized for sustaining the rate required by
real-time processing in LHCb, but can be used as a more general GPU data
processing as it includes a scheduler and a memory manager that can be
integrated into Gaudi. This permits the extension and implementation of
Allen within other experimental software using the same underlying
framework.

@@

Heterogeneous computing is still a rapidly developing and changing
field. The burden of writing architecture-specific code repeatedly will
certainly exceed the available effort from the experiments. Efforts are
ongoing in this direction, since the CWP, trying to take advantage of
the latest development in synergy with industry and attempting to
increase the use of centers with heterogeneous resources by experiments.

>>>>

Experiments have also been working towards minimising the differences
between trigger (online) and offline software for reconstruction and
calibration, so that more refined physics objects can be obtained
directly within the HLT farm for a more efficient event selection. This
is also in-line with enhancing the experiment's capabilities for
real-time data analysis of events accepted by the hardware trigger
system, driven by use cases where the physics potential improves when
analysing more events than what can be written out in full with
traditional data processing.

@@

The huge increase in the number of charged particles and hence the
combinatorial load for tracking algorithms at future colliders will put
great strain on compute resources. To minimise the memory footprint,
tracking software needs to be thread-safe to support multi-threaded
execution per core. At the same time, the software has to be efficient
and accurate to meet the physics requirements.

@@

The aim of the \textbf{mkFit} project is to speed up Kalman filter (KF)
tracking algorithms using highly parallel architectures and to deliver
track building (and possibly fitting) software for the HL-LHC. Recent
activities focused on delivering a production quality software setup to
perform track building in the context of the LHC Run-3 (or Run-2). The
implementation relies on single-precision floating point mathematics and is
available for multicore CPUs and achieves significant gains in compute
performance from the use of vector instructions set extending to
AVX-512, based on Matriplex library (a part of the mkFIt project).

\textbf{Exa.TrkX}~\cite{Ju:2020xty} is a cross-experiment collaboration of data scientists
and computational physicists from ATLAS, CMS and DUNE. It develops
production-quality deep neural network models, in particular Graph
Neural Networks, for charged particle tracking on diverse detectors
employing next-generation computing architectures such as HPCs. It is
also exploring distributed training and optimisation of Graph Neural
Networks (GNN) on HPCs, and the deployment of GNNs with microsecond
latencies on Level-1 trigger systems.

@@

\hypertarget{general-reconstruction-software-improvements-vectorisation}{%
\subsubsection{General reconstruction software improvements:
vectorisation}\label{general-reconstruction-software-improvements-vectorization}}

Improving the ability of HEP developed toolkits to use vector units on
commodity hardware will bring speedups to applications running on both
current and future hardware. The goal for work in this area is to evolve
current toolkit and algorithm implementations, and best programming
techniques to better use SIMD capabilities of current and future
computing architectures. Since the CWP, algorithm development projects
(e.g. \cite{LHCB-FIGURE-2019-002}
as well as mkFit) have demonstrated success in increasing the use of vector
units. In addition, HEP has increasingly benefited from industry
developed software, including machine learning toolkits (e.g.,
TensorFlow), that typically make excellent use of vector units.
Challenges remain in this area. For example, full exploitation of this
must also account for many generations of hardware that experiments must
exploit on the grid. In addition, to realise full performance gains, a
large fraction of the application must be improved to use SIMD processor
capabilities (partly due to turbo-boost capabilities of processors)

@@

Work is already undergoing in the collaborations to evolve or rewrite
existing toolkits and algorithms focused on their physics and technical
performance at high event complexity (e.g. high pileup at HL-LHC), and
efforts in this area have expanded. Cross-collaboration developments are
focusing on the availability of ML algorithms for experimental software,
especially in persistifying models and running inferences in production
and in enhancing the
training capacities of collaborations by submitting jobs to the grid and
to facilities offering large CPU/GPU resources~\cite{LWTNN, ONNX}. One can also foresee
that HEP will increase the benefit from techniques and cross-talk from
industry.

@@

\textbf{Allen} is a data processing framework for GPUs, as well as a
specific implementation of a first-level trigger for LHCb Run 3
data-taking. Allen is optimised for sustaining the rate required by
real-time processing in LHCb, but can be used as a more general GPU data
processing as it includes a scheduler and a memory manager that can be
integrated into Gaudi. This permits the extension and implementation of
Allen within other experimental software using the same underlying
framework.

@@

Heterogeneous computing is still a rapidly developing and changing
field. The burden of writing architecture-specific code repeatedly will
certainly exceed the available effort from the experiments. Efforts are
ongoing in this direction, since the CWP, trying to take advantage of
the latest development in synergy with industry and attempting to
increase the use of centres with heterogeneous resources by experiments.
